---
title: "Individual Fairness Model"
author: "Simon Schrauth"
date: "`r Sys.Date()`"
output: html_document
---
# Individual

## Goal

After executing this script, the individual fairness/discrimination of the base model should be determined.

## Preparation
### Install packages
```{r}
# when running the script the first time
webshot::install_phantomjs()
```


```{r}
pacman::p_load(tidyverse, 
               here,
               tidymodels,
               cluster,
               scales,
               MLmetrics,
               wesanderson,
               knitr,
               kableExtra,
               magick)

tidymodels_prefer()
```

### Load data
```{r}
data = readRDS(here("01_data", "02_data_processed", "01_base_model",
                    "03_data.rds"))

model_fit = readRDS(file = here("01_data", "02_data_processed", "01_base_model",
                                "04_model_fit.rds"))
```

### Source Function Script
```{r}
source(file = here("02_scripts", "02_individual", "00_individual_functions.R"))
```

## Model Calculation
### Estimate model with custom optimization algorithm
```{r}
data_prep = recipe(compas ~ .,
                         data = data) %>% 
  step_dummy(all_nominal_predictors()) %>% 
  prep() %>% 
  bake(new_data = NULL) %>% 
  mutate(compas = as.numeric(compas)-1)
```


```{r}
model_base = glm(compas ~ ., 
                 data = data_prep, 
                 family = binomial(link = "logit"),
                 method = my_method)

model_base_summary = data.frame(estimate = model_base$coefficients,
                                se = as.vector(model_base$se),
                                z = model_base$z,
                                p_value = model_base$p)

model_base_summary_nice = model_base_summary %>% 
    slice(-1) %>% 
    rownames_to_column() %>% 
    mutate(rowname = case_when(
      rowname == "age" ~ "Alter",
      rowname == "days_in_jail" ~ "Tage im Gefängnis",
      rowname == "juv_fel_count" ~ "Straftaten (jugendl.)",
      rowname == "juv_misd_count" ~ "Ordnungswidr. (jugendl.)",
      rowname == "juv_other_count" ~ "Andere Anzeigen (jugendl.)",
      rowname == "priors_count" ~ "Anzahl Straftaten",
      rowname == "sex_Female" ~ "Geschlecht: Weiblich",
      rowname == "race_African.American" ~ "Race: Schwarz",
      rowname == "race_Hispanic" ~ "Race: Hispanisch",
      rowname == "race_Other" ~ "Race: Andere",
      rowname == "charge_degree_M" ~ "Schwere der Straftat: Ordnungswidr.",
      rowname == "two_year_recid_yes" ~ "Rückfälligkeit: Ja",
      .default = rowname
        )
      ) %>% 
    column_to_rownames(var = "rowname") %>% 
    mutate(estimate = exp(estimate)) %>% 
    mutate(estimate_percent = percent((estimate - 1), digits = 1)) %>% 
    relocate(estimate_percent, .after = estimate) %>% 
    rename("Schätzer" = estimate,
           "Schätzer (in Prozent)" = estimate_percent,
           Standardfehler = se,
           "z-Wert" = z,
           "p-Wert" = p_value)

kable(model_base_summary_nice, booktabs = TRUE, 
      align = "c", 
      digits = c(3, 1, 3, 2, 4)) %>% 
  row_spec(seq(1, nrow(model_base_summary_nice), 2), background = "#f8f8f8") %>% 
  kable_styling(html_font = "serif") %>% 
  save_kable(file = here("03_outputs", "02_individual", "08_table_base.png"),
             density = 500)
```

### Calculate individual discrimination
```{r}
# extract probabilities out of base model
base_prob = my_predict(summary = model_base_summary, 
                       data = data_prep)

# calculate proportion of cases where individual fairness is not given (D(Mx, My) > d(x,y))
dist_base = diff_dist(prob = base_prob, 
                      data = data)

dist_base$n_unfair_rel

plot_base_rel = data.frame(x = c("Fair", "Unfair"),
                           y = c(1-dist_base$n_unfair_rel, dist_base$n_unfair_rel)) %>% 
  ggplot(aes(x = x, y = y, fill = factor(1))) +
    geom_col() +
    geom_text(aes(label = percent(y, accuracy = 0.1)), vjust = -0.35, family = "serif") +
    labs(title = "Anteil an fairen Paar-Vergleichen",
         subtitle = expression("mit individueller Fairness als D("*plain("M"[x])*","*plain("M"[y])*") \u2264 d(x,y)"),
         x = "",
         y = "Anteil an allen Paar-Vergleichen") +
    scale_fill_manual(values = wes_palette("Darjeeling1")[5]) +
    scale_y_continuous(labels = percent) +
    guides(fill = "none") +
    theme(text = element_text(family = "serif"))
```

### Save output
```{r}
saveRDS(model_base,
        file = here("01_data", "02_data_processed", "02_individual", 
                    "01_model_base.rds"))

saveRDS(dist_base,
        file = here("01_data", "02_data_processed", "02_individual", "big_data",
                    "01_dist_base.rds"))
```

```{r}
model_base = readRDS(file = here("01_data", "02_data_processed", "02_individual", 
                                 "01_model_base.rds"))

dist_base = readRDS(file = here("01_data", "02_data_processed", "02_individual", "big_data",
                                "01_dist_base.rds"))
```

```{r}
plot_base_matrix = dist_base$diff_dist %>% 
  as.data.frame() %>%
  rename_with(.fn = ~ str_replace(., "V", "")) %>% 
  rename_with(.fn = ~ str_replace(., "^(\\d)$", "000\\1")) %>% 
  rename_with(.fn = ~ str_replace(., "^(\\d{2})$", "00\\1")) %>% 
  rename_with(.fn = ~ str_replace(., "^(\\d{3})$", "0\\1")) %>% 
  rownames_to_column() %>% 
  mutate(rowname = str_replace(rowname, "^(\\d)$", "000\\1")) %>% 
  mutate(rowname = str_replace(rowname, "^(\\d{2})$", "00\\1")) %>% 
  mutate(rowname = str_replace(rowname, "^(\\d{3})$", "0\\1")) %>% 
  pivot_longer(cols = !c(rowname)) %>% 
  ggplot(aes(x = name, y = rowname, fill = factor(value))) +
    geom_tile() +
    labs(title = "Individuelle Fairness für jeden Paar-Vergleich aller Daten-Punkte",
         subtitle = "für Orginial-Modell",
         x = "",
         y = "") +
  scale_fill_manual(name = "", 
                    labels = c("Fair", "Unfair", "x"), 
                    values = c("0" = wes_palette("Darjeeling1")[2], 
                               "1" = wes_palette("Darjeeling1")[1],
                               "NA" = "gray")) +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        text = element_text(family = "serif"))
```

```{r}
diff_dist_num = dist_base$diff_dist_num
diff_dist_num[upper.tri(diff_dist_num)] = t(diff_dist_num)[upper.tri(diff_dist_num)]

diff_dist_ordered = diff_dist_num %>% 
  as.data.frame() %>% 
  summarise(across(.cols = everything(),
                   .fns = ~sum(.x, na.rm = TRUE))) %>% 
  pivot_longer(cols = everything()) %>%
  mutate(value = value/2) %>% 
  arrange(desc(value))

plot_base_dens = diff_dist_ordered %>% 
  ggplot() +
    geom_vline(xintercept = 1, linetype = "dashed", color = "grey") +
    geom_histogram(aes(x = value, y = after_stat(density)),
                   bins = 50,
                   fill = wes_palette("Darjeeling1")[5],
                   alpha = 0.5) +
    geom_density(aes(x = value),
                 linewidth = 1,
                 color = wes_palette("Darjeeling1")[5]) +
    labs(title = "Differenz der Input- und Output-Unterschiede aller Daten-Paare",
         subtitle = expression("Histogramm und Dichte für D("*plain("M"[x])*","*plain("M"[y])*")-d(x,y)"),
         x = expression("Differenz von D("*plain("M"[x])*","*plain("M"[y])*") und d(x,y)"),
         y = "Dichte") +
    theme(text = element_text(family = "serif"))
```


```{r}
ggsave("01_plot_base_rel.png",
       plot = plot_base_rel,
       device = "png",
       path = here("03_outputs", "02_individual"),
       dpi = 500,
       width = 20,
       height = 12,
       units = c("cm")
       )

ggsave("02_plot_base_matrix.png",
       plot = plot_base_matrix,
       device = "png",
       path = here("03_outputs", "02_individual"),
       dpi = 500,
       width = 30,
       height = 30,
       units = c("cm")
       )

ggsave("03_plot_base_dens.png",
       plot = plot_base_dens,
       device = "png",
       path = here("03_outputs", "02_individual"),
       dpi = 500,
       width = 20,
       height = 12,
       units = c("cm")
       )
```


## In-Processing
### Hyper-Parameter Tuning (lambda)
```{r}
lambdas = seq(-120, 0, 20)

results_fair = c()
results_acc = c()

for (lambda in lambdas) {
  print(paste("Lambda:", lambda))
  
  model_temp = glm(compas ~ ., 
                   data = data_prep, 
                   family = binomial(link = "logit"),
                   method = my_method_with_penalty,
                   control = list(lambda = lambda,
                                  penalty_function = my_penalty_function))

  model_temp_summary = data.frame(estimate = model_temp$coefficients,
                                  se = as.vector(model_temp$se),
                                  z = model_temp$z,
                                  p_value = model_temp$p)
  
  pen_prob = my_predict(summary = model_temp_summary, 
                        data = data_prep)
  

  dist_linf_pen = diff_dist(prob = pen_prob, 
                            data = data)
  
  logloss = LogLoss(pen_prob$low, data_prep$compas)
  
  results_fair = append(results_fair, values = dist_linf_pen$n_unfair_rel)
  results_acc = append(results_acc, values = logloss)
}
```

```{r}
hyperparam_results = data.frame(Lambda = lambdas,
                                Fairness_Metrik = results_fair,
                                Log_Loss = results_acc) %>% 
  mutate(Kosten_Funktion = Log_Loss + Lambda * Fairness_Metrik)

lambdas = seq(-120, 0, 20)

plot_hyperparam_results = hyperparam_results %>% 
  pivot_longer(cols = !Lambda) %>% 
  mutate(name = str_replace(name, "_", "-")) %>% 
  ggplot(aes(x = Lambda, y = value, color = name)) +
    geom_vline(xintercept = -100,
               color = wes_palette("Darjeeling1")[3]) +
    geom_point() +
    geom_line() +
    facet_wrap(~factor(name, levels = c("Log-Loss",
                                        "Fairness-Metrik",
                                        "Kosten-Funktion")
                       ), 
               scales = "free_y") +
    scale_color_manual(values = wes_palette("Darjeeling1")[c(2,1,5)]) +
    scale_x_continuous(breaks = lambdas) +
    labs(title = "Hyperparameter-Tuning für \u03bb",
         subtitle = "Log-loss, Fairness-Metrik & Kosten-Funktion (Log-loss + \u03bb \u00d7 Fairness-Metrik) \nbei unterschiedlichen \u03bb-Werten",
         y = "") +
    guides(color = "none") +
    theme(text = element_text(family = "serif"))
```


```{r}
saveRDS(hyperparam_results,
        file = here("01_data", "02_data_processed", "02_individual", 
                    "03_hyperparam_results.rds"))

```

```{r}
hyperparam_results = readRDS(file = here("01_data", "02_data_processed", "02_individual", 
                                         "03_hyperparam_results.rds"))
```

```{r}
ggsave("04_plot_hyperparam_results.png",
       plot = plot_hyperparam_results,
       device = "png",
       path = here("03_outputs", "02_individual"),
       dpi = 500,
       width = 20,
       height = 12,
       units = c("cm")
       )
```


### Final fair model
```{r}
final_lambda = -100

model_fair = glm(compas ~ ., 
                 data = data_prep, 
                 family = binomial(link = "logit"),
                 method = my_method_with_penalty,
                 control = list(lambda = final_lambda, 
                                penalty_function = my_penalty_function))

model_fair_summary = data.frame(estimate = model_fair$coefficients,
                                se = as.vector(model_fair$se),
                                z = model_fair$z,
                                p_value = model_fair$p)

model_fair_summary %>% 
    slice(-1) %>% 
    mutate(estimate = exp(estimate)) %>% 
    mutate(estimate_percent = percent((estimate - 1), digits = 1)) %>% 
    relocate(estimate_percent, .after = estimate)

model_fair_summary_nice = model_fair_summary %>% 
    slice(-1) %>% 
    rownames_to_column() %>% 
    mutate(rowname = case_when(
      rowname == "age" ~ "Alter",
      rowname == "days_in_jail" ~ "Tage im Gefängnis",
      rowname == "juv_fel_count" ~ "Straftaten (jugendl.)",
      rowname == "juv_misd_count" ~ "Ordnungswidr. (jugendl.)",
      rowname == "juv_other_count" ~ "Andere Anzeigen (jugendl.)",
      rowname == "priors_count" ~ "Anzahl Straftaten",
      rowname == "sex_Female" ~ "Geschlecht: Weiblich",
      rowname == "race_African.American" ~ "Race: Schwarz",
      rowname == "race_Hispanic" ~ "Race: Hispanisch",
      rowname == "race_Other" ~ "Race: Andere",
      rowname == "charge_degree_M" ~ "Schwere der Straftat: Ordnungswidr.",
      rowname == "two_year_recid_yes" ~ "Rückfälligkeit: Ja",
      .default = rowname
        )
      ) %>% 
    column_to_rownames(var = "rowname") %>% 
    mutate(estimate = exp(estimate)) %>% 
    mutate(estimate_percent = percent(round((estimate - 1),3), digits = 1)) %>% 
    relocate(estimate_percent, .after = estimate) %>% 
    rename("Schätzer" = estimate,
           "Schätzer (in Prozent)" = estimate_percent,
           Standardfehler = se,
           "z-Wert" = z,
           "p-Wert" = p_value)

kable(model_fair_summary_nice, booktabs = TRUE, 
      align = "c", 
      digits = c(3, 1, 3, 2, 4)) %>% 
  row_spec(seq(1, nrow(model_fair_summary_nice), 2), background = "#f8f8f8") %>% 
  kable_styling(html_font = "serif") %>% 
  save_kable(file = here("03_outputs", "02_individual", "09_table_fair.png"),
             density = 500)
```

Calculate individual discrimination
```{r}
# extract probabilities out of base model
prob_fair = my_predict(summary = model_fair_summary, 
                       data = data_prep)

# calculate proportion of cases where individual fairness is not given (D(Mx, My) > d(x,y))
dist_fair = diff_dist(prob = prob_fair, 
                      data = data)

data_plot_fair_rel = data.frame(x = c("Fair", "Unfair"),
                                y = c(1-dist_fair$n_unfair_rel, dist_fair$n_unfair_rel),
                                data = "In-processed Modell")

plot_fair_rel = plot_base_rel$data %>% 
  mutate(data = "Original-Modell") %>% 
  rbind(data_plot_fair_rel) %>% 
  mutate(data = factor(data, levels = c("Original-Modell", "In-processed Modell"))) %>% 
  ggplot(aes(x = x, y = y, fill = data)) +
    geom_col(position = position_dodge(0.82), width = 0.8) +
    geom_text(aes(label = percent(y, accuracy = 0.1)), 
              position = position_dodge(0.82), 
              vjust = -0.35,
              family = "serif") +
    labs(title = "Anteil an fairen Paar-Vergleichen",
         subtitle = expression("mit individueller Fairness als D("*plain("M"[x])*","*plain("M"[y])*") \u2264 d(x,y)"),
         x = "",
         y = "Anteil an allen Paar-Vergleichen",
         fill = "") +
    scale_fill_manual(values = wes_palette("Darjeeling1")[c(5,3)]) +
    scale_y_continuous(labels = percent) +
    theme(text = element_text(family = "serif"),
          legend.position = "top")
```


```{r}
plot_fair_matrix = dist_fair$diff_dist %>% 
  as.data.frame() %>%
  rename_with(.fn = ~ str_replace(., "V", "")) %>% 
  rename_with(.fn = ~ str_replace(., "^(\\d)$", "000\\1")) %>% 
  rename_with(.fn = ~ str_replace(., "^(\\d{2})$", "00\\1")) %>% 
  rename_with(.fn = ~ str_replace(., "^(\\d{3})$", "0\\1")) %>% 
  rownames_to_column() %>% 
  mutate(rowname = str_replace(rowname, "^(\\d)$", "000\\1")) %>% 
  mutate(rowname = str_replace(rowname, "^(\\d{2})$", "00\\1")) %>% 
  mutate(rowname = str_replace(rowname, "^(\\d{3})$", "0\\1")) %>% 
  pivot_longer(cols = !c(rowname)) %>% 
  ggplot(aes(x = name, y = rowname, fill = factor(value))) +
    geom_tile() +
    labs(title = "Individuelle Fairness für jeden Paar-Vergleich aller Daten-Punkte",
         subtitle = "für In-processed Modell",
         x = "",
         y = "") +
  scale_fill_manual(name = "", 
                    labels = c("Fair", "Unfair", "x"), 
                    values = c("0" = wes_palette("Darjeeling1")[2], 
                               "1" = wes_palette("Darjeeling1")[1],
                               "NA" = "gray")) +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        text = element_text(family = "serif"))
```

```{r}
diff_dist_num_fair = dist_fair$diff_dist_num
diff_dist_num_fair[upper.tri(diff_dist_num_fair)] = t(diff_dist_num_fair)[upper.tri(diff_dist_num_fair)]

diff_dist_ordered_fair = diff_dist_num_fair %>% 
  as.data.frame() %>% 
  summarise(across(.cols = everything(),
                   .fns = ~sum(.x, na.rm = TRUE))) %>% 
  pivot_longer(cols = everything()) %>%
  mutate(value = value/2) %>% 
  arrange(desc(value))

plot_fair_dens = plot_base_dens +
    geom_histogram(data = diff_dist_ordered_fair,
                   aes(x = value, y = after_stat(density)),
                   bins = 50,
                   fill = wes_palette("Darjeeling1")[3],
                   alpha = 0.5) +
    geom_density(data = diff_dist_ordered_fair,
                 aes(x = value),
                 linewidth = 1,
                 color = wes_palette("Darjeeling1")[3]) +
    labs(title = "Differenz der Input- und Output-Unterschiede aller Daten-Paare",
         subtitle = expression("Histogramm und Dichte für D("*plain("M"[x])*","*plain("M"[y])*")-d(x,y)"),
         x = expression("Differenz von D("*plain("M"[x])*","*plain("M"[y])*") und d(x,y)"),
         y = "Dichte") +
    theme(text = element_text(family = "serif")) +
    geom_label(x = -820, y = 0.0035, 
               label = "In-processed", 
               color = wes_palette("Darjeeling1")[3],
               family = "serif") +
    geom_label(x = -50, y = 0.002, 
               label = "Original", 
               color = wes_palette("Darjeeling1")[5],
               family = "serif") 
```


Save output
```{r}
saveRDS(model_fair,
        file = here("01_data", "02_data_processed", "02_individual", 
                    "02_model_fair.rds"))

saveRDS(dist_fair,
        file = here("01_data", "02_data_processed", "02_individual", "big_data",
                    "02_dist_fair.rds"))
```

```{r}
model_fair = readRDS(file = here("01_data", "02_data_processed", "02_individual", 
                                 "02_model_fair.rds"))

dist_fair = readRDS(file = here("01_data", "02_data_processed", "02_individual", "big_data",
                                "02_dist_fair.rds"))
```

```{r}
ggsave("05_plot_fair_rel.png",
       plot = plot_fair_rel,
       device = "png",
       path = here("03_outputs", "02_individual"),
       dpi = 500,
       width = 20,
       height = 12,
       units = c("cm")
       )

ggsave("06_plot_fair_matrix.png",
       plot = plot_fair_matrix,
       device = "png",
       path = here("03_outputs", "02_individual"),
       dpi = 500,
       width = 30,
       height = 30,
       units = c("cm")
       )

ggsave("07_plot_fair_dens.png",
       plot = plot_fair_dens,
       device = "png",
       path = here("03_outputs", "02_individual"),
       dpi = 500,
       width = 20,
       height = 12,
       units = c("cm")
       )
```

