---
title: "Counterfactual Fairness"
author: "Simon Schrauth"
date: "`r Sys.Date()`"
output: html_document
---

# Counterfactual Fairness

## Goal

After executing this script, the counterfactual fairness/discrimination of the base model should be determined.

## Install packages

Installing rstan package (and the dependent StanHeaders package) from source 
(for additional information or in the case of installation problems see also https://github.com/stan-dev/rstan/wiki/RStan-Getting-Started)

```{r}
install.packages("StanHeaders", repos = c("https://mc-stan.org/r-packages/", getOption("repos")))
install.packages("rstan", repos = c("https://mc-stan.org/r-packages/", getOption("repos")))
```


```{r}
pacman::p_load(tidyverse, 
               here,
               fastDummies,
               dagitty,
               ggdag,
               lavaan,
               rstan,
               wesanderson)

options(mc.cores = parallel::detectCores())
rstan_options(auto_write = TRUE)
```

## Load data

```{r}
data_train = readRDS(here("01_data", "02_data_processed",
                          "04_data_train.rds"))
```

## Prepare data
```{r}
data_counter = data_train %>% 
  select(Age = age,
         Sex = sex,
         Race = race,
         JuvFel = juv_fel_count,
         JuvMisd = juv_misd_count,
         Priors = priors_count,
         Crime = charge_degree,
         COMPAS = compas) %>% 
  dummy_cols(select_columns = c("Race", "Sex"),
             remove_first_dummy = FALSE,
             remove_selected_columns = TRUE) %>% 
  rename("Race_Native_American" = "Race_Native American",
         "Race_African_American" = "Race_African-American") %>% 
  mutate(Crime = ifelse(Crime == "F", 1, 0),
         COMPAS = ifelse(COMPAS == "High", 1, 0)) 
```


## Build DAG (directed acyclic graph)
```{r}
dag = dagitty('dag {
  Age [exposure,pos="0,-2"]
  Sex [pos="0,-1"]
  Race [pos="0,0"]
  Priors [pos="0.5, -1.75"]
  Crime [pos="1, -1.75"]
  U_D [latent,pos="2, -1.5"]
  JuvMisd [pos="0.5, -0.25"]
  JuvFel [pos="1,-0.25"]
  U_J [latent,pos="2, -0.5"]
  COMPAS [outcome,pos="1.5,-1"]
  Race -> Priors
  Race -> Crime
  Race -> JuvMisd
  Race -> JuvFel
  Race -> COMPAS
  Sex -> Priors
  Sex -> Crime
  Sex -> JuvMisd
  Sex -> JuvFel
  Sex -> COMPAS
  Age -> Priors
  Age -> Crime
  Age -> JuvMisd
  Age -> JuvFel
  Age -> COMPAS
  U_D -> Priors
  U_D -> Crime
  U_J -> JuvMisd
  U_J -> JuvFel
  U_D -> COMPAS
  }')

colors = wes_palette("Darjeeling1")

dag_plot = tidy_dagitty(dag) %>% 
  mutate(name = str_replace(name, "U_J", "U[J]")) %>% 
  mutate(name = str_replace(name, "U_D", "U[D]")) %>% 
  ggplot(aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_dag_edges(data_directed = function(x) subset(x, name != "U[J]" & name != "U[D]")) +
  geom_dag_edges(data_directed = function(x) subset(x, to == "COMPAS"),
                 edge_color = colors[3]) +
  geom_dag_edges_arc(data = function(x) subset(x, name == "U[J]"),
                          curvature = -1) +
  geom_dag_edges_arc(data = function(x) subset(x, name == "U[D]" & to != "COMPAS"),
                          curvature = 1) +
  geom_dag_node(data = function(x) subset(x, name %in% c("U[J]", "U[D]")),
                 color = colors[2]) +
  geom_dag_node(data = function(x) subset(x, name %in% c("Race", "Sex", "Age")),
                 color = colors[5]) +
  geom_dag_node(data = function(x) subset(x, name=="COMPAS"),
                 color = colors[3]) +
  geom_dag_node(data = function(x) subset(x, name %in% c("Priors", 
                                                         "Crime",
                                                         "JuvMisd",
                                                         "JuvFel")),
                 color = colors[4]) +
  geom_dag_text(size = 2.2, parse = TRUE) +
  theme_dag()

dag_plot

tidy_dagitty(dag)
```

```{r}
ggsave("XX_dag.png",
       plot = dag_plot,
       device = "png",
       path = here("03_outputs", "03_outputs_graphs"),
       dpi = 300,
       width = 25,
       height = 15,
       units = c("cm")
       )
```


## Build alternative DAG (directed acyclic graph)
```{r}
dag2 = dagitty('dag {
  Age [exposure,pos="0,-2"]
  Sex [pos="0,-1"]
  Race [pos="0,0"]
  Priors [pos="0.5, -1.5"]
  Crime [pos="1, -1.5"]
  U_D [latent,pos="2, -1.5"]
  JuvMisd [pos="0.5, -0.5"]
  JuvFel [pos="1,-0.5"]
  U_J [latent,pos="2, -0.5"]
  COMPAS [outcome,pos="1.5,-1"]
  Race -> Priors
  Race -> Crime
  Race -> JuvMisd
  Race -> JuvFel
  Race -> COMPAS
  Sex -> Priors
  Sex -> Crime
  Sex -> JuvMisd
  Sex -> JuvFel
  Sex -> COMPAS
  Age -> Priors
  Age -> Crime
  Age -> JuvMisd
  Age -> JuvFel
  Age -> COMPAS
  U_D -> Priors
  U_D -> Crime
  U_J -> JuvMisd
  U_J -> JuvFel
  U_D -> COMPAS
  Priors -> COMPAS
  Crime -> COMPAS
  JuvMisd -> COMPAS
  JuvFel -> COMPAS
  }')

colors = wes_palette("Darjeeling1")

dag2_plot = tidy_dagitty(dag2) %>% 
  mutate(name = str_replace(name, "U_J", "U[J]")) %>% 
  mutate(name = str_replace(name, "U_D", "U[D]")) %>% 
  ggplot(aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_dag_edges(data_directed = function(x) subset(x, name != "U[J]" & name != "U[D]")) +
  geom_dag_edges(data_directed = function(x) subset(x, to == "COMPAS"),
                 edge_color = colors[3]) +
  geom_dag_edges_arc(data = function(x) subset(x, name == "U[J]"),
                          curvature = -1) +
  geom_dag_edges_arc(data = function(x) subset(x, name == "U[D]" & to != "COMPAS"),
                          curvature = 1) +
  geom_dag_node(data = function(x) subset(x, name %in% c("U[J]", "U[D]")),
                 color = colors[2]) +
  geom_dag_node(data = function(x) subset(x, name %in% c("Race", "Sex", "Age")),
                 color = colors[5]) +
  geom_dag_node(data = function(x) subset(x, name=="COMPAS"),
                 color = colors[3]) +
  geom_dag_node(data = function(x) subset(x, name %in% c("Priors", 
                                                         "Crime",
                                                         "JuvMisd",
                                                         "JuvFel")),
                 color = colors[4]) +
  geom_dag_text(size = 2.2, parse = TRUE) +
  theme_dag()

dag2_plot

tidy_dagitty(dag2)
```
```{r}
ggsave("XX_dag2.png",
       plot = dag2_plot,
       device = "png",
       path = here("03_outputs", "03_outputs_graphs"),
       dpi = 300,
       width = 25,
       height = 15,
       units = c("cm")
       )
```


## Build Structural Causal Model via Bayesian Estimation
## with RStan (adapted code from https://github.com/mkusner/counterfactual-fairness)

# Prepare data for modelling with Stan
```{r}
data_counter_protected = data_counter %>% 
  select(starts_with(c("Race_", "Sex_"))) 

n = nrow(data_counter)

data_counter_list = list(N = n,
                         K = length(data_counter_protected),
                         prot = data.matrix(data_counter_protected),
                         age = data_counter[, "Age", drop = TRUE],
                         priors = data_counter[, "Priors", drop = TRUE],
                         crime = data_counter[, "Crime", drop = TRUE],
                         juvmisd = data_counter[, "JuvMisd", drop = TRUE],
                         juvfel = data_counter[, "JuvFel", drop = TRUE],
                         compas = data_counter[, "COMPAS", drop = TRUE])
```

# Fit model
```{r}
fit_counter <- stan(file = here("02_scripts", "02_discrimination_identification",
                                "04_counterfactual_fairness_stan.stan"), 
                    data = data_counter_list, 
                    iter = 2000, 
                    chains = 4,
                    verbose = TRUE)
```

# Save model
```{r}
saveRDS(fit_counter, 
        file = here("01_data", "02_data_processed",
                    "XX_model_fit_counter.rds")
        )
```

```{r}
model_counter = extract(fit_counter)

uj_est = colMeans(model_counter$u_j)
mean(model_counter$compas_ud)

```

```{r}
saveRDS(model_counter, 
        file = here("01_data", "02_data_processed",
                    "XX_model_counter.rds")
        )
```