---
title: "Data Cleaning"
author: "Simon Schrauth"
date: "`r Sys.Date()`"
output: html_document
---

# Clean Compas Dataset

## Goal

After executing this script, the compas dataset should be cleaned and ready for modelling. Additionally, one should have an overview over the variables and their missingness pattern.

## Install packages

```{r}
pacman::p_load(tidyverse, 
               here,
               naniar,
               lubridate,
               wesanderson,
               ggdist)
```

## Load data

```{r}
data_raw = readRDS(here("01_data", "02_data_processed", "01_base_model",
                        "01_data_recid_load.rds"))
```

## Glimpse on the data

### Overview over data

```{r}
data_raw
```

```{r}
str(data_raw)
```

### Check for Missing Values

Table of absolute missing values per variable as well as percentage of missing values per variable

```{r}
miss_var_summary(data_raw)
```

Plot with most frequent missingness patterns with according absolute frequencies

```{r}
gg_miss_upset(data_raw) 
```

## Cleaning

INSERT TEXT HERE

```{r}
data_clean = data_raw %>% 
  arrange(id) %>% 
  column_to_rownames(var = "id") %>% 
  select(sex, 
         race,
         age,
         juv_fel_count,
         juv_misd_count,
         juv_other_count,
         priors_count = priors_count...15, 
         charge_degree = c_charge_degree, 
         days_b_screening_arrest, 
         is_recid, 
         c_jail_in, 
         c_jail_out,
         two_year_recid,
         compas = score_text) %>% 
  filter(days_b_screening_arrest <= 30,
         days_b_screening_arrest >= -30,
         is_recid != -1,
         charge_degree != "O") %>% 
  mutate(compas = factor(ifelse(compas == "Low", 1, 0))) %>% 
  mutate(c_jail_in = floor_date(c_jail_in, unit = "day"),
         c_jail_out = floor_date(c_jail_out, unit = "day")) %>% 
  mutate(days_in_jail = as.numeric(round(difftime(c_jail_out, 
                                                  c_jail_in, 
                                                  unit = "days")
                                         )
                                   )
         ) %>% 
  mutate(across(.cols = c(sex,
                          race,
                          charge_degree),
                .fns = ~ factor(.))) %>% 
  mutate(sex = fct_relevel(sex, "Male", after = 0)) %>% 
  mutate(race = fct_relevel(race, "Caucasian", after = 0)) %>% 
  mutate(two_year_recid = factor(two_year_recid, 
                                 levels = c("0", "1"),
                                 labels = c("no", "yes"))) %>% 
  select(!c(days_b_screening_arrest,
            c_jail_in,
            c_jail_out,
            is_recid)) %>%
  select(sex,
         race,
         age,
         days_in_jail,
         juv_fel_count,
         juv_misd_count,
         juv_other_count,
         priors_count,
         charge_degree,
         two_year_recid,
         compas)
```


## Exploratory Data Analysis
### Dataset overview
```{r}
data_clean
```

```{r}
str(data_clean)
```

### Visualization of the class variable (readmitted) & the protected variables (race, age & gender)

#### Definition of colors
```{r}
colors_discrete = rep(wes_palette("Darjeeling1"), 3)
```

#### Class variable: Recidivism
```{r}
eda_class = data_clean %>% 
  select(compas) %>% 
  ggplot(aes(x = factor(compas), fill = factor(compas))) +
  geom_bar() +
  labs(title = "Häufigkeit der Klassen-Variable im COMPAS-Datensatz",
       subtitle = "mit Klassen-Variable 'COMPAS-Score'",
       x = "COMPAS-Score", 
       y = "Häufigkeit") +
  scale_x_discrete(labels = c("Hohes Risiko", "Niedriges Risiko")) +
  scale_fill_manual(values = colors_discrete) +
  guides(fill = "none")

eda_class
```

```{r}
ggsave("01_eda_class.png",
       plot = eda_class,
       device = "png",
       path = here("03_outputs", "01_base_model"),
       dpi = 300,
       width = 25,
       height = 15,
       units = c("cm")
       )
```

#### Protected variable: Race
```{r}
race_relabel = function(x){
  x = str_replace(x, "African-American", "Schwarz")
  x = str_replace(x, "Caucasian", "Weiß")
  x = str_replace(x, "Hispanic", "Hispanisch")
  x = str_replace(x, "Other", "Andere")
  x = str_replace(x, "Asian", "Asiatisch")
  x = str_replace(x, "Native American", "Amerikanisch-Indigen")
  
  return(x)
}

eda_race = data_clean %>% 
  select(race) %>% 
  mutate(race = fct_relabel(.f = race, .fun = ~ race_relabel(.x))) %>% 
  ggplot(aes(x = fct_infreq(race), fill = fct_infreq(race))) +
  geom_bar() +
  labs(title = "Häufigkeit der Race-Variable im COMPAS-Datensatz",
       x = "Race", 
       y = "Häufigkeit") +
  scale_fill_manual(values = colors_discrete[c(1,2,4,5,3,2)]) +
  guides(fill = "none")

eda_race
```

```{r}
ggsave("02_eda_race.png",
       plot = eda_race,
       device = "png",
       path = here("03_outputs", "01_base_model"),
       dpi = 300,
       width = 25,
       height = 15,
       units = c("cm")
       )
```

#### Protected variable: Age
```{r}
eda_age_raincloud = data_clean %>% 
  select(age) %>% 
  ggplot(aes(x = age, fill = factor(1))) +
  ggdist::stat_halfeye(
    adjust = 0.75,
    justification = -.2,
    .width = 0,
    point_colour = NA
  ) +
  geom_boxplot(
    width = .12,
    outlier.colour = NA,
    alpha = 0.5
  ) +
  labs(title =  "Alters-Verteilung im COMPAS-Datensatz",
       subtitle = "Wahrscheinlichkeitsdichte-Verteilung & Boxplot",
       x = "Alter", 
       y = "") +
  scale_fill_manual(values = colors_discrete[3]) +
  guides(fill = "none")

eda_age_raincloud
```

```{r}
eda_age_hist = data_clean %>% 
  select(age) %>% 
  ggplot(aes(x = age, fill = factor(1))) +
  geom_histogram(binwidth = 2) +
  labs(title = "Alters-Verteilung im COMPAS-Datensatz",
       subtitle = "Histogramm mit Balkenbreite 2",
       x = "Alter", 
       y = "Dichte") +
  scale_fill_manual(values = colors_discrete[3]) +
  guides(fill = "none") 

eda_age_hist
```

```{r}
ggsave("03_eda_age_raincloud.png",
       plot = eda_age_raincloud,
       device = "png",
       path = here("03_outputs", "01_base_model"),
       dpi = 300,
       width = 25,
       height = 15,
       units = c("cm")
       )

ggsave("04_eda_age_hist.png",
       plot = eda_age_hist,
       device = "png",
       path = here("03_outputs", "01_base_model"),
       dpi = 300,
       width = 25,
       height = 15,
       units = c("cm")
       )
```

#### Protected variable: Gender
```{r}
eda_sex = data_clean %>% 
  select(sex) %>% 
  mutate(sex = ifelse(sex == "Male", "Männlich", "Weiblich")) %>% 
  ggplot(aes(x = fct_infreq(sex), fill = fct_infreq(sex))) +
  geom_bar() +
  labs(title =  "Geschlechter-Verteilung im COMPAS-Datensatz",
       x = "Geschlecht", 
       y = "Häufigkeit") +
  scale_fill_manual(values = colors_discrete[c(2,3)]) +
  guides(fill = "none")

eda_sex
```

```{r}
ggsave("05_eda_sex.png",
       plot = eda_sex,
       device = "png",
       path = here("03_outputs", "01_base_model"),
       dpi = 300,
       width = 15,
       height = 15,
       units = c("cm")
       )
```


#### Protected variables: Race & Gender
```{r}
eda_race_sex = data_clean %>% 
  count(race, sex) %>% 
  mutate(race = fct_relabel(.f = race, .fun = ~ race_relabel(.x))) %>% 
  mutate(sex = ifelse(sex == "Male", "Männlich", "Weiblich")) %>% 
  group_by(race) %>% 
  mutate(n_cum = sum(n)) %>% 
  ungroup() %>% 
  mutate(n_rel = n/n_cum) %>% 
  ggplot(aes(x = sex, y  = n_rel, fill = sex)) +
  geom_col() +
  facet_wrap(~race) +
  labs(title = "Geschlechter-Verteilung im COMPAS-Datensatz",
       subtitle = "gruppiert nach Race",
       x = "Geschlecht", 
       y = "Häufigkeit") +
  scale_fill_manual(values = colors_discrete[c(2,3)]) +
  guides(fill = "none") +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  scale_y_continuous(labels = scales::percent)

eda_race_sex
```

```{r}
ggsave("06_eda_race_sex.png",
       plot = eda_race_sex,
       device = "png",
       path = here("03_outputs", "01_base_model"),
       dpi = 300,
       width = 25,
       height = 15,
       units = c("cm")
       )
```



#### Protected variables: Race & Age
```{r}
eda_race_age = data_clean %>% 
  select(age, race) %>% 
  mutate(race = fct_relabel(.f = race, .fun = ~ race_relabel(.x))) %>%
  mutate(race = factor(race, levels = c("Andere",
                                        "Amerikanisch-Indigen",
                                        "Asiatisch",
                                        "Hispanisch",
                                        "Schwarz", 
                                        "Weiß"
                                        )
                       )
         ) %>% 
  ggplot(aes(x = race, y = age, fill = race)) +
  ggdist::stat_halfeye(
    adjust = 0.75,
    justification = -.05,
    .width = 0,
    point_colour = NA
  ) +
  geom_boxplot(
    width = .12,
    outlier.colour = NA,
    alpha = 0.5
  ) +
  labs(title = "Alters-Verteilung im COMPAS-Datensatz",
       subtitle = "gruppiert nach Race",
       x = "Race", 
       y = "Alter") +
  scale_fill_manual(values = colors_discrete[c(1,2,4,5,3,2)]) +
  guides(fill = "none") +
  coord_flip()

eda_race_age
```

```{r}
ggsave("07_eda_race_age.png",
       plot = eda_race_age,
       device = "png",
       path = here("03_outputs", "01_base_model"),
       dpi = 300,
       width = 25,
       height = 15,
       units = c("cm")
       )
```


#### Protected variables: Age & Gender
```{r}
eda_sex_age = data_clean %>% 
  select(age, sex) %>% 
  mutate(sex = ifelse(sex == "Male", "Männlich", "Weiblich")) %>% 
  mutate(sex = factor(sex, levels = c("Weiblich", "Männlich"))) %>% 
  ggplot(aes(x = sex, y = age, fill = sex)) +
  ggdist::stat_halfeye(
    adjust = 0.75,
    justification = -.05,
    .width = 0,
    point_colour = NA
  ) +
  geom_boxplot(
    width = .12,
    outlier.colour = NA,
    alpha = 0.5
  ) +
  labs(title = "Alters-Verteilung im COMPAS-Datensatz",
       subtitle = "gruppiert nach Geschlecht",
       x = "Geschlecht", 
       y = "Alter") +
  scale_fill_manual(values = colors_discrete[c(3,2)]) +
  guides(fill = "none") +
  coord_flip()

eda_sex_age
```

```{r}
ggsave("08_eda_sex_age.png",
       plot = eda_sex_age,
       device = "png",
       path = here("03_outputs", "01_base_model"),
       dpi = 300,
       width = 25,
       height = 15,
       units = c("cm")
       )
```

## Final Data Cleaning with respect to results from EDA
```{r}
data = data_clean %>% 
  mutate(race = factor(case_when(
    race == "Asian" ~ "Other",
    race == "Native American" ~ "Other",
    .default = race
                                )
                      )
         ) %>% 
  mutate(race = fct_relevel(race, "Caucasian", after = 0))
```


## Save data

```{r}
saveRDS(data, 
        file = here("01_data", "02_data_processed", "01_base_model",
                    "03_data.rds")
        )
```



